Simple repression draft

### 2020-05-16 doc/paper.tex
Finished discussion of means + other small edits

\mmnote{Snappy names for all these cartoons would really help the reader. Also naming them by number of promoter states is confusing to the reader because analogous models have different number of states when we're talking about constitutive version vs version with repressor. Thoughts??}

@@

In fact it was noted in~\cite{Chure2019} that this two-state
model can be viewed as the coarse-graining of any equilibrium
promoter model in which no transcriptionally active states have
transcription factor bound. Put differently, there is no overlap
between TF bound states and transcriptionally active states. We
will see this explicitly in the 3-state equilibrium model below,
but perhaps surprising is that something analogous carries over
even to the nonequilbrium models we consider later.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
\caption{
\textbf{
An overview of the simple repression motif at the level of means.}
(A) depicts the qualitative biological picture of simple repression.
(B) shows a variety of possible mathematized cartoons of simple
repression, along with the corresponding complexity $\rho$ of the
architecture, an effective parameter which subsumes all details
of the architecture excluding the repressor.
(C) shows the ``master curve'' to which all cartoons in (B) collapse.
}
\label{fig1:means_cartoons}
\end{figure}

@@

\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
Analogous to the three-state equilibrium model, this model
fine-grains the repressor unbound state of nonequilbrium model 1,
resolving it into an empty promoter state and an RNAP-bound state.
Note in this picture, in contrast with model 4 below,
transcription initiation is accompanied by a promoter state
change, in keeping with the interpretation as RNAP-bound and
empty states: if an RNAP successfully escapes the promoter and
proceeds to elongation of a transcript, clearly it is no longer
bound at the promoter! Therefore another RNAP must bind before
another transcript can be initiated.

@@

\subsubsection{Nonequilibrium model 3
- Multistep transcription initiation and escape}
One might reasonably complain that the first two
``nonequilbrium'' models we have considered are straw men. Their
steady states necessarily satisfy detailed balance which is
equivalent to thermodynamic equilibrium. Why is this the case? At
steady state there is by definition no net probability flux in or
out of each promoter state, but since the promoter states form a
linear chain, there is only one way in or out of the repressor
bound and RNAP bound states, implying each edge must actually
have a net zero probability flux, which is the definition of
detailed balance (usually phrased as equality of forward and
reverse transition fluxes).

@@

\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
\mmnote{Key ideas:
\begin{itemize}
\item Exclusive repression and occupancy hypothesis. Not
necessarily steric hindrance, e.g., maybe kinked DNA, but it is
key that there is zero transcription when repressor is bound.
This is the key takeaway from the qualitative cartoon in Fig 1A.
\item Key assumption: expression is proportional to occupation
probability of whatever the ``active'' state is for a given
model, whatever we mean by active.
\item Also emphasize that parameters/rates are \textit{not}
apples-to-apples between cartoons, especially the initiation rate $r$.
Indeed, this is the entire point! Except for repressor
energies/rates, the other parameters are not independently
measurable or inferrable b/c we're not even sure what
steps/states we should be writing in the model!
(simple RNAP kinetics as inferred in vitro, or supercoiling, or
RNAP cluster kinetics a la Ibrahim, or maybe
gene-copy/RNAP-copy/other-copy number is really the key, or...)
So all you get are coarse-grained combos like $\rho$, and even
$\rho$ would be hard to infer, much less trying to infer its innards.
\item Emphasize that tuning any of the parameters in $\rho$ is
probably off-limits experimentally: it wrecks entire cell
physiology and confounds everything. You can write down models of
what you THINK the important steps are and aren't, and use in
vitro data for the important steps, but without any way to
experimentally perturb in vivo, you have no way to TEST whether
your modeling assumptions are correct! Also point out
supercoiling~\cite{Chong2014,Sevier2016} to emphasize that the
potential features that are necessary to include are WAY more
than just RNAP details. One possible tuning knob is maybe RNAP
binding sites, but we don't have a terrific predictive
understanding of how mutations there would influence any of the
parameters inside $\rho$, so still not obvious how to make use of
that. Brewster/Daniel2012 energy matrix is only good in one model
framework and doesn't tell you anything about how model params in
other pictures would respond.
In theory you could refit their raw data to different models, but
as we see later, even having data of full mRNA distributions is
insufficient to identify parameters in all but the simplest of
the models in Figs.\ 1 and 2, so it seems doubtful that only
measurements of means would fare better.
??Double check, what did they actually measure in 2012 paper??
\item Weak promoter corresponds to $\rho\approx 1$. In principle,
if repressor binding E or kinetics are known to high enough
precision, deviations from $\rho=1$ become a matter of expt. In
practice, I don't think it's at all realistic to measure
repressor params or expression levels precisely enough to detect
the expected tiny deviations from $\rho=1$. We can estimate: how
good would determination of $\Delta F_R$ in Fig~1B have to be to
get a given uncertainty in $\rho$? I think it's approx 1-to-1:
e.g., an uncertainty of 0.1 (kT) in $\Delta F_R$ converts to a
0.1 uncert in $\log\rho$, and since $\rho\approx 1$, that means
$\log(1+\Delta\rho)\approx\Delta\rho\approx 0.1$.
That's a really hard determination of $\Delta F_R$ for a very
coarse determination of $\rho$. Doubtful that inference of
kinetic rates would be any easier and probably a lot harder.Also,
in vitro RNAP data says $\rho\approx 1$ should be true, and even
if that's the wrong quantity to be talking about, Manuel's
inference of $k^+$ and $k^-$ suggests that still $\rho$ is close
to 1, and that inference is agnostic to the microscopics.
\item Master curve exists b/c repressor is on its own leaf of
graph. (But maybe don't use that language.) Arb noneq stuff can
happen on rest of promoter state space, doesn't matter. Is this
``one curve to rule them all'' a tautology? How do we probe the
theory? Can't do it at level of means, there's no discriminatory
power, need to go beyond. Test if free energy inferred from
fold-change measurements agrees with TF rates inferred from
single molec, and if those rates agree with rates inferred from
mRNA population distributions.
\item Here's a fun puzzle: 2-state equilibrium model gets details
of transcription so wrong, how can it be useful? Because
fold-change is a ratio! Since details all collapse into $\rho$,
fitting this 2-state model to fold-change measurements can
\textit{still} give you a not-bad estimate of ratio of repressor
rates, i.e., binding energies.
\item 
\end{itemize}

>>>>

@@

In fact it was noted in~\cite{Chure2019} that this two-state
model can be viewed as the coarse-graining of any equilibrium
promoter model in which no transcriptionally active states have
transcription factor bound. Put differently, there is no overlap
between TF bound states and transcriptionally active states. We
will see this explicitly in the 3-state equilibrium model below,
but perhaps surprising is that something analogous carries over
even to the nonequilibrium models we consider later.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/fig1/fig1.ai}
\caption{
\textbf{
An overview of the simple repression motif at the level of means.}
(A) depicts the qualitative biological picture of simple repression.
(B) shows a variety of possible mathematized cartoons of simple
repression, along with the effective parameter $\rho$
which subsumes all regulatory details
of the architecture excluding the repressor.
(C) shows the ``master curve'' to which all cartoons in (B) collapse.
}
\label{fig1:means_cartoons}
\end{figure}

@@

\subsubsection{Nonequilibrium model 2 - RNAP bound/unbound states}
Analogous to the three-state equilibrium model, this model
fine-grains the repressor unbound state of nonequilibrium model 1,
resolving it into an empty promoter state and an RNAP-bound state.
Note in this picture, in contrast with model 4 below,
transcription initiation is accompanied by a promoter state
change, in keeping with the interpretation as RNAP-bound and
empty states: if an RNAP successfully escapes the promoter and
proceeds to elongation of a transcript, clearly it is no longer
bound at the promoter! Therefore another RNAP must bind before
another transcript can be initiated.

@@

\subsubsection{Nonequilibrium model 3
- Multistep transcription initiation and escape}
One might reasonably complain that the first two
``nonequilibrium'' models we have considered are straw men. Their
steady states necessarily satisfy detailed balance which is
equivalent to thermodynamic equilibrium. Why is this the case? At
steady state there is by definition no net probability flux in or
out of each promoter state, but since the promoter states form a
linear chain, there is only one way in or out of the repressor
bound and RNAP bound states, implying each edge must actually
have a net zero probability flux, which is the definition of
detailed balance (usually phrased as equality of forward and
reverse transition fluxes).

@@

\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
What have we learned from our exploration of the models
in~\fig{fig1:means_cartoons} and their predictions of
fold-changes in mean expression levels? The takeaway is the
existence of a master curve (\fig{fig1:means_cartoons}C) to which
the fold-change predictions of all the models collapse. This
master curve is parametrized by only two effective parameters:
$\Delta F_R$, which characterizes the repressor, and $\rho$,
which characterizes all other features of the promoter
architecture. The key assumption underpinning this result is that
zero transcription occurs when a repressor is bound to its
operator, though note we are agnostic about the molecular
mechanism which achieves this.\footnote{
Steric effects are the obvious mechanism, but, for instance,
``action-at-a-distance'' mediated by kinked DNA due to
repressors bound tens or hundreds of nucleotides upstream of a
promoter is plausible as well.}

Why does the master curve of~\fig{fig1:means_cartoons}C exist at all?
This brings to mind the deep questions posed in,
e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search
for general ``symmetries'' of the problem instead of specifying
generative models of the known microscopics of a system.
In our case, the key feature seems to be the exclusive nature of
repressor and RNAP binding, which allows the parameter describing
the repressor, $\Delta F_R$, to cleanly separate from all other
details of the promoter architecture, which are encapsulated in $\rho$.
In the graph-theoretic language which was applied to similar
problems in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the
repressor-bound state is a leaf of the graph describing
transitions between promoter states. Arbitrary nonequilibrium
behavior can occur on the rest of the promoter state space, but
it may all be swept up in the effective parameter $\rho$, to
which the repressor makes no contribution.

As suggested in~\cite{Chure2019}, we believe this master curve
should generalize to architectures with multiple repressor
binding sites, as long as the exclusivity of transcription factor
binding and transcription initation is maintained. The
interpretation of $\Delta F_R$ is then of an effective free
energy of all repressor bound states.
In an equilibrium picture this is simply given by the log of the
sum of Boltzmann weights of all repressor bound states, which
looks like the log of a partition function of a subsystem.
In a nonequilibrium picture, while we can still mathematically gather
terms and give the resulting collection the label $\Delta F_R$,
it is unclear if the physical interpretation as an effective free energy
makes sense, and we leave this an open problem for future work.\footnote{
As is well known, free energies cannot be assigned unambiguously
to states out of equilibrium because the free energy change along
a generic path traversing the state space is path dependent,
unlike at equilibrium.
A consequence of this is that, out of equilibrium, $\Delta F_R$
is no longer a simple sum of Boltzmann weights. Instead it
resembles a restricted sum of King-Altman
diagrams~\cite{King1956, Hill1966}. Following the work of
Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
as an effective free energy, but this remains unclear to us.
}

If we relax the requirement of exclusive repressor-RNAP binding,
one could imagine models in which repressor and RNAP doubly-bound
states are allowed, where the repressor's effect is to
\text{reduce} the transcription rate rather than setting it to zero.
Our results do not strictly apply to such a model, although we
note that if the repressor's reduction of the transcription rate
is substantial, such a model might still be well-approximated by
one of the models in~\fig{fig1:means_cartoons}.

One may worry that our ``one curve to rule them all'' is a
mathematical tautology. In fact we \textit{agree} with this
criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
be meaningfully interpreted as a real, physical free energy.
We believe our theory deserves to be scrapped unless it can be
demonstrated that $\Delta F_R$ \textit{is} a physically
meaningful and interpretable free energy.

How do we test this and probe the theory with fold-change
measurements? There is a fundamental limitation in that the
master curve is essentially a one-parameter function of $\Delta
F_R + \log\rho$. Worse, there are many \textit{a priori}
plausible microscopic mechanisms that could contribute to the
value of $\rho$, such as RNAP binding and escape
kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
more speculatively, RNAP clusters analogous to those obvserved in
eukaryotes~\cite{Cisse2013}. Even if $\Delta F_R$ is measured to
high precision, inferring the potential microscopic contributions
to $\rho$, buried inside a log no less, from fold-change
measurements seems hopeless. As a statistical inference problem
it is entirely nonidentifiable, in the language
of~\cite{Gelman2013}, Section 4.3.

If we cannot simply infer values of $\rho$ from measurements of
fold-change, can we perturb some of the parameters that make up
$\rho$ and measure the change? Unfortunately this is probably
off-limits experimentally: most of the potential contributors to
$\rho$ are global processes that affect many or all genes. For
instance, changing RNAP association rates by changing RNAP copy
numbers, or changing supercoiling kinetics by changing
topoisomerase copy numbers, would massively perturb the entire
cell's physiology and confound any determination of $\rho$.

One might instead imagine a bottom-up modeling approach, where we
mathematize a model of what we hypothesize the important steps
are and are not, use in vitro data for the steps deemed
important, and \textit{predict} what $\rho$ should be.
But again, because of the one-parameter nature of the master
curve, many different models will likely make indistinguishable
predictions, and without any way to experimentally perturb in
vivo, there is no clear way to test whether the modeling
assumptions are correct.

In light of this, we prefer the view that parameters and rates
are not directly comparable between cartoons
in~\fig{fig1:means_cartoons}. Rather, parameters in the
simpler cartoons represent coarse-grained combinations of
parameters in the finer-grained models. For instance, by equating
$\rho$ between any two models, one can derive various possible
correspondences between the two models' parameters. Note that
these correspondences are clearly not unique, since many possible
associations could be made. It then is a choice as to what microscopic
interpretations the model-builder prefers for the parameters in a
particular cartoon, and as to which coarse-grainings lend
intuition and which seem nonsensical. Indeed, since it remains an open question what microscopic features dominate $\rho$ (as suggested above, perhaps RNAP binding and escape
kinetics~\cite{DeHaseth1998, Mitarai2015}, or supercoiling
accumulation and release~\cite{Chong2014, Sevier2016}, or,
something more exotic like RNAP clusters~\cite{Cisse2013}), we are hesitant to put too much weight on any one microscopic interpretation of model parameters that make up $\rho$.

One possible tuning knob to probe $\rho$ that would not globally
perturb the cell's physiology is to manipulate RNAP binding sites.
Work such as~\cite{Kinney2010} has shown that models of
sequence-dependent RNAP can be inferred from data, and the
authors of~\cite{Brewster2012} showed that the model
of~\cite{Kinney2010} has predictive power by using the model to
\textit{design} binding sites of a desired affinity.
But for our purposes, this begs the question: the authors
of~\cite{Kinney2010} \textit{assumed} a particular model
(essentially our 3-state equilibrium model but without the
repressor), so it is unclear how or if such methods can be turned
around to \textit{compare} different models of promoter function.

We have already pointed out that the master curve
of~\fig{fig1:means_cartoons} is essentially a one-parameter model,
the one parameter being $\Delta F_R + \log\rho$.
By now the reader may be alarmed as to how can we even determine
$\Delta F_R$ and $\rho$ independently of each other, never mind
shedding a lens on the internal structure of $\rho$ itself.
A hint is provided by the weak promoter approximation, invoked
repeatedly in prior studies~\cite{Bintu2005c, Garcia2011a,
Razo-Mejia2018} of simple repression using the 3-state
equilibrium model in~\fig{fig1:means_cartoons}.
In that picture, the weak promoter approximation means
$\frac{P}{N_{NS}}\exp(-\beta\Delta\epsilon_P) \ll 1$,
meaning therefore $\rho\approx1$.
This is suggestive, but how can we be sure that $\rho$ is not,
for instance, actually $10^2$ and that $\Delta F_R$ hides a
compensatory factor?
A resolution is offered by an independent inference of $\rho$ in
the absence of repressors. This was done in~\cite{Razo-Mejia2020}
by fitting nonequilibrium model 4  in~\fig{fig1:means_cartoons},
with zero repressor (looking ahead, this is equivalent to model 4
in~\fig{fig:constit_cartoons}), to FISH data
from~\cite{Brewster2014}. This provided a determination of $k^+$
and $k^-$, from which their ratio is estimated to be no more than
a few $10^{-1}$ and possibly as small as $10^{-2}$.

The realization that $\rho\approx1$ to an excellent
approximation, \textit{independent} of which model
in~\ref{fig1:means_cartoons} one prefers, goes a long way towards
explaining the surprising success of equilibrium models of simple
repression. Even though our 2- and 3-state models get so many
details of transcription wrong, it does not matter because
fold-change is a cleverly designed ratio. Since $\rho$ subsumes
all details except the repressor, and $\log\rho\approx0$, fitting
these simple models to fold-change measurements can still give a
surprisingly good estimate of repressor binding energies. So the
ratiometric construction of fold-change fulfills its intended
purpose of canceling out all features of the promoter
architecture except the repressor itself. Nevertheless it is
perhaps surprising how effectively it does so: \textit{a priori},
one might not have expected $\rho$ to be quite so close to 1.

We would also like to highlight the relevance
of~\cite{Landman2019} here. Landman et.\ al.\ reanalyzed and
compared in vivo and in vitro data on the lacI repressor's
binding affinity to its various operator sequences. (The in vivo
data was from, essentially, fitting our master curve to
expression measurements.) They find broad agreement between the
in vitro and in vivo values. This reinforces the suspicion that
the equilibrium $\Delta\epsilon_R$ repressor binding energies do
in fact represent real physical free energies.
Again, \textit{a priori} this did not have to be the case, even
knowing that $\rho\approx1$.

In principle, if $\Delta F_R$ can be measured to sufficient
precision, then deviations from $\rho=1$ become a
testable matter of experiment.
In practice, it is probably unrealistic to measure repressor
rates $k_R^+$ or $k_R^-$ or fold-changes in expression levels
(and hence $\Delta\epsilon_R$) precisely enough to detect the
expected tiny deviations from $\rho=1$.
We can estimate the
requisite precision in $\Delta F_R$ to resolve a given
$\Delta\rho$ by noting, since $\rho\approx1$, that
$\log(1+\Delta\rho)\approx \Delta\rho$, so
$\Delta(\Delta F_R) \approx \Delta\rho$.
Suppose we are lucky and $\Delta\rho$ happens to be $\sim0.1$, on
the high end of our range estimated above.
A determination of $\Delta\epsilon_R/k_BT$ with an uncertainty of
barely 0.1 was achieved in the excellent measurements
of~\cite{Razo-Mejia2018}, so this requires a very difficult
determination of $\Delta F_R$ for a very crude determination of $\rho$,
which suggests, to put it lightly, this is not a
promising path to pursue experimentally.
It is doubtful that inference of repressor kinetic rates
would be any easier.\footnote{
\mmnote{
In fact the uncertainties quoted in Hammar et.\
al.~\cite{Hammar2014} from Elf's group are small enough to do
this calculation, but HJ \& I think their error bars are insanely
optimistic. Not sure I want to blatantly call them out on it though??}
}

Moving forward, we have weak evidence supporting the
interpretation of $\Delta F_R$ as a physically real free
energy~\cite{Landman2019} and other work casting
doubt~\cite{Hammar2014}.
How might we resolve the confusion?
If there is no discriminatory power to test the theory and
distinguish the various models with measurements of fold-changes
in means, how do we probe the theory? Clearly to discriminate
between the nonequilibrium models in~\fig{fig1:means_cartoons},
we need to go beyond means to ask questions about kinetics, noise and even
full distributions of mRNA copy numbers over a population of cells.
If the ``one-curve-to-rule-them-all'' is more than a
mathematical tautology, then the free energy of repressor binding
inferred from fold-change measurements should agree with
repressor binding and unbinding rates.
In other words, the equilbrium and nonequilibrium definitions of
$\Delta F_R$ must agree, meaning
\begin{equation}
\Delta F_R = \beta\Delta\epsilon_R - \log(R/N_{NS})
        = - \log(k_R^+/k_R^-),
\end{equation}
must hold, where $\beta\Delta\epsilon_R$ is inferred from the
master curve fit to fold-change measurements, and $k_R^+$ and
$k_R^-$ are inferred in some orthogonal manner.
Single molecule measurements such as from~\cite{Hammar2014} have
directly observed these rates, and in the remainder of this work
we explore a complementary approach: inferring repressor binding
and unbinding rates $k_R^+$ and $k_R^-$ from FISH measurements of
mRNA population distributions.
